Optimizing a PoS Tag Set for Dependency Parsing

### 2016-05-05 submission.tex
Fixes to experimental setup

Similarly, \newcite{Mac:05} investigated the effects of PoS tag sets on tagger
performance in English, specifically the Wall Street Journal portion of Penn
Treebank \cite{Mar:San:Mar:93}. They mapped the original tag set of Penn
Treebank to more fine-grained tag sets using linguistic insight to investigate
whether additional linguistic information included in finer-grained tags could
assist the tagger. Experimenting with both lexically and syntactically
conditioned modifications such as distinguishing between count nouns and
noncount nouns and between transitive and intransitive verbs, they found that
more fine-grained tag sets rarely led to improvements in tagger accuracy; the
most successful modification yielded an improvement in tagger accuracy of 0.05
percentage points.

@@

\paragraph{Morphological Annotation}
The morphological annotation and PoS tag set of NDT follows the Oslo-Bergen
Tagger \cite{Hag:Joh:Nok:00,Sol:13}, which in turn is largely based on the work
of \newcite{Faa:Lie:Van:97}. The tag set consists of 12 morphosyntactic PoS
tags, with 7 additional tags for punctuation and symbols. The tag set is thus
rather coarse-grained, with broad categories such as \texttt{subst} (noun) and
\texttt{verb} (verb). The PoS tags are complemented by a large set of
morphological features, providing information about morphological properties such as
definiteness, number and tense. These features are used in our tag set
modifications, where the coarse PoS tag of relevant tokens is concatenated with
one or more of these features to include more linguistic information in the
tags.

\begin{table}
    \centering
    %\smaller[0.5]
    \begin{tabular}{@{}ll@{}}
        \toprule
        \textbf{Tag} & \textbf{Description} \\
        \midrule
        \texttt{adj} & Adjective \\
        \texttt{adv} & Adverb \\
        \texttt{det} & Determiner\\
        \texttt{inf-merke} & Infinitive marker \\
        \texttt{interj} & Interjection \\
        \texttt{konj} & Conjunction \\
        \texttt{prep} & Preposition \\
        \texttt{pron} & Pronoun \\
        \texttt{sbu} & Subordinate conjunction \\
        \texttt{subst} & Noun \\
        \texttt{ukjent} & Unknown (foreign word) \\
        \texttt{verb} & Verb \\
        \bottomrule
    \end{tabular}
    \caption{Overview of the PoS tag set of NDT.}
    \label{ndttagest}
\end{table}

@@

Our approach is somewhat semi-automatic in that we use a "mix" of both manual
annotation in our experiments. Our initial tags are gold standard, and using
linguistic insight coupled with computational considerations, we introduce new
features or feature combination to the tag set. These tags are run through a
pipeline with automatically assigned tags in both training and testing, where
the most promising modifications are used "further". Hence, the tag set
modifications are not deterministic, nor do we try all possible combination; we
only experiment with linguistically motivated distinctions.

\paragraph{Data Set Split}
As there was no standardized data set split of NDT due to its very recent
development, we needed to establish a data set split
(training/development/test). Our data set split of the treebank follows the
standard 80-10-10 (training/development/test) split and will be distributed
with the treebank and proposed as the new standard(?). In creating the data set
split, care has been taken to preserve contiguous texts in the various data
sets while keeping the split balanced in terms of genre (and source). Our
proposed data set split was used in the Norwegian contribution to the Universal
Dependencies project \cite{Ovr:Hoh:16}.  The split will be made available at a
companion website.

\paragraph{Tagger}
For our experiments with tag set modifications, we wanted a PoS tagger both
reasonably fast and accurate. There is often a trade-off between the two
factors/considerations, as the most accurate taggers tend to suffer in terms of
speed due to their complexity. However, a tagger that achieves both close to
state-of-the-art accuracy as well as very high speed is TnT \cite{Bra:00}. TnT
was furthermore recently used to evaluate the recently proposed universal tag
set \cite{Pet:Das:McD:12}. The sum of these factors led to TnT being the tagger
of choice for our experiments.

\paragraph{Parser}
In choosing a syntactic parser for our experiments, we considered previous work
on dependency parsing of Norwegian, specifically that of \cite{Sol:Skj:Ovr:14}.
They found the Mate parser \cite{Boh:10} to be the most successful parser for
the parsing of NDT. Furthermore, recent dependency parser comparisons
\cite{Cho:Tet:Ste:15} showed that Mate performed very well on parsing of
English, beating a range of contemporary state-of-the-art parsers.

\paragraph{Tag Set Mapping}
In order to carry out the tag set modifications, we created a mapping that
takes as input a set of tags and associated morphological features to be
included in said tags. This mapping maps the relevant existing tags to new,
more fine-grained tags including more relevant morphological features for the
applicable tokens. This is done by concatenating the original tag with said
features for all applicable tokens, i.e., tokens that are assigned said tag and
contains the set of features in its morphological features.

\paragraph{Baseline}
It is common practice to compare the performance of PoS taggers to a
pre-computed baseline for an initial point of comparison.
For PoS tagging, a commonly used baseline is the Most Frequent Tag (MFT)
baseline, which we use in our experiments. This involves labeling each word
with the tag it was assigned most frequently in the training. All unknown
words, i.e., words not seen in the training data, are assigned the tag most
frequently assigned to words seen only once in the training. Unknown and
infrequent words have in common that they rarely occur, and we might therefore
expect them to have similar properties.

\paragraph{Tags \& Features}
As we seek to quantify the effects of PoS tagging in a realistic setting, i.e.,
in application to raw text, we want to evaluate the parser on automatically
assigned PoS tags. For the training of the parser, however, we have two
options: using either gold standard or automatically assigned tags. In order to
settle on a configuration, we conducted experiments with gold standard and
automatically assigned tags to see how they differ with respect to performance.
The results of our experiments, shown in Table \ref{tagconfigeval}, reveal that
the combination of training and testing on automatic tags is clearly superior
to training on gold standard tags and testing on automatic tags. Consequently,
the parser was both trained and tested on automatically assigned tags in our
experiments.

@@

We removed any morphological features in order to simulate a realistic setting.
Moreover, it is crucial that we remove these features when working with
automatically assigned tags, as the features is still gold standard.  For
instance, if a verb token is erroneously tagged as a noun, we could potentially
have a noun token with verbal features such as tense, which markedly obfuscates
the training and parsing. Another important aspect is that we want to isolate
the effect of PoS tags, necessitating the exclusion of morphological features.

@@

\section{Tag Set Optimization}
\label{sec:optimization}
By introducing more fine-grained linguistically motivated distinctions in a tag
set, we increase the linguistic information represented in the tags, which may
assist the parser in recognizing and generalizing syntactic patterns. However,
the addition of more linguistic information to the tags and thus a more
fine-grained tag set will most likely lead to a drop in tagger accuracy/tagging
quality due to the increase in complexity. The best tagging does not
necessarily lead to the best parse, and it is therefore interesting to
investigate how the tag set modifications may affect the interplay between
tagging and parsing.

As we want to investigate how we can improve the
linguistic quality of a tag set, we will not introduce tag set modifications
which are not linguistically motivated. We will only consider distinctions we
deem linguistically sensible, even if we expect them to impair the
computational tractability; investigating the interplay between these two
considerations is ultimately our goal.

\subsection{Baseline Experiments}
In an initial round of experiments, we concatenated the tag of each token with
its set of morphological features in order to map the original tag set to a
new, more fine-grained tag set (hereafter referred to as the \emph{full} tag
set). This resulted in a total of 368 tags, which is clearly very fine-grained.
The two initial tag sets, i.e., the original tag set comprising 19 tags and the
full tag set comprising 368 tags, thus represent two extremes in terms of
granularity.  To evaluate these tag sets and investigate how the tag set
granularity would affect the performance of tagging and parsing, we trained and
evaluated TnT and Mate on the training and development data, respectively, on
the two tag sets. In Table \ref{inittagseteval}, we report the results of these
experiments. We see that the tagger accuracy drastically drops when going from
the original to the full tag set. TnT reports an accuracy of 97.47\% on the
original tag set, which is reduced to 93.48\% for the full tag set. These
results confirm our hypothesis that the very high linguistic quality in the
full, fine-grained tag set comes at the expense of drops in tagger performance.
However, the additional linguistic information provided by the full tag set
improves the parser performance. With the original tag set, Mate reports an LAS
of 87.01\% and a UAS of 90.19\%, which increases to 87.15\% and 90.39\%,
respectively, when using the full tag set. As we are looking for linguistically
informed distinctions that improve the syntactic parsing, these results are
promising and serve to indicate that additional morphological information
assists the syntactic parsers, motivating the optimization of the existing PoS
tag set.

@@

@@

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}llllll@{}}
        \toprule
        \textbf{Data Set} & \textbf{Tag Set} & \textbf{MFT} & \textbf{Accuracy}
        & \textbf{LAS} & \textbf{UAS} \\
        \midrule
        \multirow{2}{*}{Dev}
        & Original & \textbf{94.14\%} & \textbf{97.47\%} & 87.01\% & 90.19\% \\
        %Full & 85.15\% & 93.48\% & 87.15\% & 90.39\% \\
        & Optimized & 89.20\% & 96.85\% & \textbf{88.87\%} & \textbf{91.78\%} \\
        \midrule
        \multirow{2}{*}{Test}
        & Original & 94.22\% & 97.38\% & 86.64\% & 90.07\% \\
        %& Full & 84.27\% & 92.79\% & 87.03\% & 90.16\% \\
        & Optimized & 88.08\% & 96.35\% & 88.55\% & 91.41\% \\
        \bottomrule
    \end{tabular}
    \caption{Results of tagging and parsing with the optimized tag set,
        compared to the initial tag sets. Trained and tested using
        automatically assigned tags from TnT.}
    \label{finalresults}
\end{table}

>>>>

Similarly, \newcite{Mac:05} investigated the effects of PoS tag sets on tagger
performance in English, specifically the Wall Street Journal portion of Penn
Treebank \cite{Mar:San:Mar:93}. They mapped the original tag set of Penn
Treebank comprising 45 tags to more fine-grained tag sets using linguistic
insight to investigate whether additional linguistic information included in
finer-grained tags could assist the tagger. Experimenting with both lexically
and syntactically conditioned modifications such as distinguishing between
count nouns and noncount nouns and between transitive and intransitive verbs,
they found that more fine-grained tag sets rarely led to improvements in tagger
accuracy; the most successful modification yielded an improvement in tagger
accuracy of 0.05 percentage points.  They did not find any statistically
significant improvements using linguistically informed distinctions, arguing
that their results do not support the hypothesis that it is possible to achieve
significant performance improvements in PoS tagging by utilizing a
finer-grained tag set.

@@

\paragraph{Morphological Annotation}
The morphological annotation and PoS tag set of NDT follows the Oslo-Bergen
Tagger \cite{Hag:Joh:Nok:00,Sol:13}, which in turn is largely based on the work
of \newcite{Faa:Lie:Van:97}. The tag set consists of 12 morphosyntactic PoS
tags outlined in Table \ref{ndttagset}, with 7 additional tags for punctuation
and symbols. The tag set is thus rather coarse-grained, with broad categories
such as \texttt{subst} (noun) and \texttt{verb} (verb). The PoS tags are
complemented by a large set of morphological features, providing information
about morphological properties such as definiteness, number and tense. These
features are used in our tag set modifications, where the coarse PoS tag of
relevant tokens is concatenated with one or more of these features to include
more linguistic information in the tags.

\begin{table}
    \centering
    %\smaller[0.5]
    \begin{tabular}{@{}ll@{}}
        \toprule
        \textbf{Tag} & \textbf{Description} \\
        \midrule
        \texttt{adj} & Adjective \\
        \texttt{adv} & Adverb \\
        \texttt{det} & Determiner\\
        \texttt{inf-merke} & Infinitive marker \\
        \texttt{interj} & Interjection \\
        \texttt{konj} & Conjunction \\
        \texttt{prep} & Preposition \\
        \texttt{pron} & Pronoun \\
        \texttt{sbu} & Subordinate conjunction \\
        \texttt{subst} & Noun \\
        \texttt{ukjent} & Unknown (foreign word) \\
        \texttt{verb} & Verb \\
        \bottomrule
    \end{tabular}
    \caption{Overview of the PoS tag set of NDT.}
    \label{ndttagset}
\end{table}

@@

Our approach is somewhat semi-automatic in that we employ a "hybrid"
combination of manual annotation and automatic labeling in our experiments. Our
initial tags are gold standard, and using linguistic insight coupled with
computational considerations, we introduce new features or feature combination
to the tag set. These tags are run through a pipeline with automatically
assigned tags in both training and testing, where the most promising
modifications are used "further". Hence, the tag set modifications are not
deterministic, nor do we try all possible combination; we only experiment with
linguistically motivated distinctions. We are in reality altering the gold
standard tags by the addition of more linguistic information.

\paragraph{Data Set Split}
As there was no standardized data set split of NDT due to its very recent
development, we needed to establish a data set split
(training/development/test). Our data set split of the treebank follows the
standard 80-10-10 (training/development/test) split and will be distributed
with the treebank and proposed as the new standard(?). In creating the data set
split, care has been taken to preserve contiguous texts in the various data
sets while keeping the split balanced in terms of genre. Our
proposed data set split was used in the Norwegian contribution to the Universal
Dependencies project \cite{Ovr:Hoh:16}.  The split will be made available at a
companion website.

\paragraph{Tagger}
For our experiments with tag set modifications, we sought a PoS tagger both
reasonably fast and accurate. There is a often a considerable trade-off between
the two factors, as the most accurate taggers tend to suffer in terms of speed
due to their complexity. However, a tagger that achieves both close to
state-of-the-art accuracy as well as very high speed is TnT \cite{Bra:00}. TnT
was furthermore recently used to evaluate the recently proposed universal tag
set \cite{Pet:Das:McD:12}. The sum of these factors led to TnT being the tagger
of choice for our experiments.

\paragraph{Parser}
In choosing a syntactic parser for our experiments, we considered previous work
on dependency parsing of Norwegian, specifically that of
\newcite{Sol:Skj:Ovr:14}, who found the Mate parser \cite{Boh:10} to be the
most successful parser for parsing of NDT. Furthermore, recent dependency
parser comparisons \cite{Cho:Tet:Ste:15} showed that Mate performed very well
on parsing of English, beating a range of contemporary state-of-the-art
parsers. Mate was consequently selected to evaluate the tag set modifications in
our experiments.

\paragraph{Tag Set Mapping}
The tag set modifications are realized by mapping the existing tags to new,
more fine-grained tags "based" on associated morphological features to be
included in specified tags. The procedure maps the relevant existing tags to
new, more fine-grained tags including more relevant morphological features for
the applicable tokens, i.e., tokens that are assigned said tag and feature(s)
in the gold standard data. Hence, we are in actuality altering the gold
standard tags by the addition of more relevant linguistic information,
producing new gold standard annotations of higher quality.

%This is done by concatenating the original tag with the
%specified features for all applicable tokens

%contains the set of features in its morphological features.

\paragraph{Baseline}
It is common practice to compare the performance of PoS taggers to a
pre-computed baseline for an initial point of comparison.  For PoS tagging, a
commonly used baseline is the Most Frequent Tag (MFT) baseline, which we used
in our experiments. This involves labeling each word with the tag it was
assigned most frequently in the training. All unknown words, i.e., words not
seen in the training data, are assigned the tag most frequently assigned to
words seen only once in the training. Unknown and infrequent words have in
common that they rarely occur, and we might therefore expect them to have
similar properties.

\paragraph{Tags \& Features}
As we seek to quantify the effects of PoS tagging in a realistic setting, i.e.,
in application to raw text, we want to evaluate the parser on automatically
assigned PoS tags. For the training of the parser, however, we have two
options: using either gold standard or automatically assigned tags. In order to
settle on a configuration, we conducted experiments with gold standard and
automatically assigned tags to see how they differ with respect to performance.
The results of these experiments, shown in Table \ref{tagconfigeval}, reveal that
the combination of training and testing on automatic tags is clearly superior
to training on gold standard tags and testing on automatic tags. Consequently,
the parser was both trained and tested on automatically assigned tags in our
experiments.

@@

We removed any morphological features in order to simulate a realistic setting.
Moreover, it is crucial that we remove these features when working with
automatically assigned tags, as the features are still gold standard. For
instance, if a verb token is erroneously tagged as a noun, we could potentially
have a noun token with verbal features such as tense, which markedly obfuscates
the training and parsing. Another important aspect is that we want to isolate
the effect of PoS tags, necessitating the exclusion of morphological features.

@@

\section{Tag Set Optimization}
\label{sec:optimization}
By introducing more fine-grained linguistically motivated distinctions in a tag
set, we increase the linguistic information represented in the tags, which may
assist the parser in recognizing and generalizing syntactic patterns. However,
the addition of more linguistic information to the tags and thus a more
fine-grained tag set will most likely lead to a drop in tagger accuracy due to
the increase in complexity. The best tagging does not necessarily lead to the
best parse, hence it is interesting to investigate how linguistically informed
tag set modifications may affect the interplay between tagging and parsing.
%As
%we want to investigate how we can improve the linguistic quality of a tag set,
%we will
%only consider tag set modifications we deem linguistically sensible.
%we will not introduce tag set modifications which are not linguistically
%motivated. We will only consider distinctions we deem linguistically sensible.

\subsection{Baseline Experiments}
In an initial round of experiments, we concatenated the tag of each token with
its set of morphological features in order to map the original tag set to a
new, more fine-grained tag set (hereafter referred to as the \emph{full} tag
set). This resulted in a total of 368 tags, which is clearly very fine-grained.
The two initial tag sets, i.e., the original tag set comprising 19 tags and the
full tag set comprising 368 tags, thus represent two extremes in terms of
granularity. To evaluate the tag sets and investigate how the tag set
granularity would affect the performance of tagging and parsing, we trained and
evaluated TnT and Mate on the training and development data, respectively, on
the two tag sets. In Table \ref{inittagseteval}, we report the results of these
experiments. We see that the tagger accuracy drastically drops when going from
the original to the full tag set. TnT reports an accuracy of 97.47\% on the
original tag set, which is reduced to 93.48\% for the full tag set. These
results confirm our hypothesis that the very high linguistic quality in the
full, fine-grained tag set comes at the expense of drops in tagger performance.
However, the additional linguistic information provided by the full tag set
improves the parser performance. With the original tag set, Mate reports an LAS
of 87.01\% and a UAS of 90.19\%, which increases to 87.15\% and 90.39\%,
respectively, when using the full tag set. As we are looking for linguistically
informed distinctions that improve the syntactic parsing, these results are
promising and serve to indicate that additional morphological information
assists the syntactic parsers, motivating the optimization of the existing PoS
tag set.

@@

The results from tagging and parsing with modifications to nouns are reported
in Table \ref{substresults}. As nouns constitute the largest class in the
treebank by far (as shown in Table \ref{baselinetagerror}), the effects are
greatest for noun tokens in terms of overall change in performance. Apart from
case, none of the tag set modifications improves the tagging. However, they all
give rise to increases in parser accuracy scores. Genitive case marks
possession, hence nouns marked with genitive case are quite different from
other nouns, taking a noun phrase as complement. Distinguishing on type is
useful and informative, as evident by the presence of separate tags for proper
and common nouns in many tag sets, such as those of PTB, SUC and UD (described
in Section \ref{sec:extagsets}). When introducing the distinction of type, we
see a large increase in parser accuracy scores, with an LAS of 88.07\% and UAS
of 91.11\%, both exceeding the baseline by more than a percentage point.
Definiteness is the most informative feature for parsing, achieving LAS of
88.27\% and UAS of 91.42\%.

@@

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}llllll@{}}
        \toprule
        \textbf{Data Set} & \textbf{Tag Set} & \textbf{MFT} & \textbf{Accuracy}
        & \textbf{LAS} & \textbf{UAS} \\
        \midrule
        \multirow{2}{*}{Dev}
        & Original & \textbf{94.14\%} & \textbf{97.47\%} & 87.01\% & 90.19\% \\
        %Full & 85.15\% & 93.48\% & 87.15\% & 90.39\% \\
        & Optimized & 89.20\% & 96.85\% & \textbf{88.87\%} & \textbf{91.78\%} \\
        \midrule
        \multirow{2}{*}{Test}
        & Original & \textbf{94.22\%} & \textbf{97.38\%} & 86.64\% & 90.07\% \\
        %& Full & 84.27\% & 92.79\% & 87.03\% & 90.16\% \\
        & Optimized & 88.08\% & 96.35\% & \textbf{88.55\%} & \textbf{91.41\%} \\
        \bottomrule
    \end{tabular}
    \caption{Results of tagging and parsing with the optimized tag set,
        compared to the initial tag sets. Trained and tested using
        automatically assigned tags from TnT.}
    \label{finalresults}
\end{table}
